# Functionals

Functionals are functions that take other functions as input and return some vector as output. Some of the most important functionals are looping functionals - efficient and clearer substitutes for for-loops.

## Readings

* [adv-r - 11 - Functionals](https://adv-r.hadley.nz/functionals.html)
    * skip 11.3.3, 11.3.4, 11.4.1, and 11.4.3
* [r4ds - 21 - Iteration](http://r4ds.had.co.nz/iteration.html)
    * 21.4 through 21.9
* [Why use purrr::map instead of lapply?](https://stackoverflow.com/questions/45101045/why-use-purrrmap-instead-of-lapply) discussion on [stackoverflow.com](https://stackoverflow.com)

## Focus

* definition of functional
* how to use looping functionals:
    * apply family: `lapply()`, `sapply()`, `vapply()`, `mapply()`, `Map()`
    * repetitious functionals: `Reduce()`, `replicate()`
    * to a lesser extent, map family from `tidyverse/purrr`: `map()`, `map_*()`, `map2()`, `pmap()`, `walk()`, etc.
* when to use looping functionals versus for-loops
* the benefits of using the apply family over the map family
* the benefits of using the map family over the apply family

## Exercises

1. A user wanted to quickly get the standard deviations of each column of `x` using `sapply`. The results, however, are nonsensicle. Inspect the user's code below and determine why the nonsense result occurred. How can the code be corrected to give what the user wanted?
    ```{r}
    x <- matrix(c(
      3, 2, 9,
      2, 4, 3,
      4, 5, 5,
      6, 3, 8
    ), ncol = 3, byrow = TRUE)
    
    sapply(x, sd)
    ```

1. Consider the following code block.
    ```{r}
    x <- c(
      1, 1, 1, 1,
      2, 2, 2, 2,
      3, 3, 3, 3,
      4, 4, 4, 4,
      5, 5, 5, 5
    )
    x <- matrix(x, nrow = 5, byrow = TRUE)
    ```
    Recreate `x` using sequences and `replicate`. 

1. ~~Consider the following code block involving matrix multiplication.~~
    ```{r}
    a <- matrix(1:4, ncol = 2)
    b <- matrix(c(5, 3, 8, 9, 3, 4), ncol = 2)
    
    a %*% b
    
    b %*% a
    ```
    ~~Using `vapply`, create a function `mm` to mimic `%*%` wihtout using `%*%`.~~

1. Consider the following matrices already contained in a list (helpful).

    ```{r}
    lst <- list(
      a <- matrix(c(5, 4, 3, 9), nrow = 2),
      b <- matrix(c(2, 9, 1, 2), nrow = 2),
      c <- matrix(c(3, 10, 4, 3), nrow = 2)
    )
    ```

    a. Recreate `lst` using a functional.
    b. Use a functional or functionals to create a $2\times 2$ matrix where each element is the average of the corresponding elements in the three matrices of `lst`. 

1. Recreate the `Map()` function without using `Map()`, `mapply()`, or any function from `purrr`.