# Functions


## Readings

* [r4ds - 19 - Functions](http://r4ds.had.co.nz/functions.html)
    * skip 19.6.2
* [adv-r - 6 - Functions](https://adv-r.hadley.nz/functions.html)
* [adv-r - 9 - Conditions](https://adv-r.hadley.nz/conditions.html)
* [here - `ifelse()` part 2][`ifelse()` part 2]
* [R Packages: A Beginner's Guide](https://www.datacamp.com/community/tutorials/r-packages-guide) by Adolfo Alvarez

## Focus

* purpose of functions
* how to write a function; basic anatomy: arguments, body
* default arguments
* dot-dot-dot (`...`) argument
* conditions: errors and warnings
* lexical scoping
* special functions
    * infix
    * replacement
* how to use `do.call()`
* packages
    * installation, loading
    * function conflicts
    
## `ifelse()` part 2

In the [previous chapter][`ifelse()` part 1], an inappropriate use of `ifelse()` was demonstrated. The success of this inappropriate use depended on lazy evaluation. With newfound knowledge of function creation, we can further explore this issue. Previously, 

```{r}
x <- c(
  7.973, 4.885,  2.676, 3.075, 7.701, 
  5.363, 4.798, -0.406, 5.814, 5.682
)

z <- numeric(length(x))

for (i in seq_along(x)) {
  ifelse(x[i] < 5, z[i] <- NA, z[i] <- x[i]^2)
}

print(z)
```

Consider the case when `i == 2`. The test `x[2] < 5` resolves to `TRUE` and the argument `z[2] <- NA` will be used. As a side-effect, the assignment `z[2] <- NA` resolves. The argument `z[2] <- x[2]^2` is never used, so it never executes. Such is the consequence of lazy evaluation.

We can, however, force arguments to evaluate non-lazily using `force()`. In the following code block, a new function `ifelse2()` is defined that forces the evaluation of all arguments. 

```{r}
x <- c(
  7.973, 4.885,  2.676, 3.075, 7.701, 
  5.363, 4.798, -0.406, 5.814, 5.682
)

ifelse2 <- function(...){
  for(x in list(...)) force(x)
  ifelse(...)
}

z2 <- numeric(length(x))

for (i in seq_along(x)) {
  ifelse2(x[i] < 5, z2[i] <- NA, z2[i] <- x[i]^2)
}

print(z2)
```

Again, the true result of `ifelse2()` and by extension `ifelse()` was never captured. The arguments, however, were evaluated in order. Thus, for all cases where `x[i] < 5 == TRUE`, `z2[i] <- NA` will occur, but then `z2[i] <- x[i]^2` will override the previous assignment. This should reinforce the idea that the inappropriate use of `ifelse()` was only successful because of lazy evaluation, not because the function was used as intended.

Note that `ifelse2()` performs the same as `ifelse()` if used correctly.

```{r}
x <- c(
  7.973, 4.885,  2.676, 3.075, 7.701, 
  5.363, 4.798, -0.406, 5.814, 5.682
)

ifelse2(x < 5, NA, x^2)
```

## Exercises

1. Consider the following code block involving matrix multiplication.
    ```{r}
    a <- matrix(1:4, ncol = 2)
    b <- matrix(c(5, 3, 8, 9, 3, 4), ncol = 2)
    
    a %*% b
    
    b %*% a
    ```
    Using for-loops, create a function `mm`, in functional form, to mimic `%*%` wihtout using `%*%`. Create the infix version of `mm` as well without copy-and-pasting any code.
    
1. Consider the following code block.
    ```{r}
    x <- c(
      1, 1, 1, 1,
      2, 2, 2, 2,
      3, 3, 3, 3,
      4, 4, 4, 4,
      5, 5, 5, 5
    )
    x <- matrix(x, nrow = 5, byrow = TRUE)
    ```
    Recreate `x` using `rep`, `list`, `do.call`, and `cbind`. Note that this is strictly an exercise and not an example of how to create matrices like `x` in practice.
    
