

# Random sampling and variate generation {#random}

```{r, include = FALSE}
set.seed(1111111)
```

In the [Introduction][Statistics review] to this book.

* obtain sample of estimators
* sample depends on "sample" of randomly generated data
* we need to know how to randomly generate data


## Pseudo-randomness

Computers cannot generate truly random numbers. Instead, complex algorithms mimic randomness to produce *pseudo-random numbers*. These algorithms begin with an arbitrary starting value or *seed* that updates with each generated number. Unless otherwise specified, the seed is often based on the internal clock of the computer's processor.

The use of an algorithm makes pseudo-randomness predictable and reproducible. If one writes a program and provides a fixed value for the seed, then the program will return the same results on repeated runs. All random number generation scripts should begin with setting the seed.

In R, setting the seed is accomplished with the `set.seed()` function that takes an arbitrary number as its argument. In the next codeblock, the `runif()` function is used to generate a random number between 0 and 1. Note that if you run this code in your own R session, the first instance of `runif()` will return a different result. When the same seed is invoked, however, the second `runif()` will return the same result as below.

```{r}
runif(n = 1)

set.seed(12479)
runif(n = 1)
```

Pending an investigation of cyclical relationships among generated values, the seed needs to be set only once. 

```{r}
# one run
set.seed(12479)

x <- runif(n = 5)

y <- runif(n = 5)

x; y

# another run that produces the same results
set.seed(12479)

x <- runif(n = 5)

y <- runif(n = 5)

x; y
```

Remember that the random number generation algorithm updates each time it is used to generate a value. This is illustrated below. The first run uses the random number generator to produce two random numbers with seed `12345`. In the subsequent run, the seed is reset to `12345` and two separate `runif()` functions generate two random numbers. The values are equivalent to the first run; the same seed was used, the same generating function was used, and a random number was requested twice. 

```{r}
# first run
set.seed(12345)

runif(n = 2)

# reset
set.seed(12345)

runif(n = 1)

runif(n = 1)
```


## Univariate random samples [INCOMPLETE]


`sample()`

For many distributions, R provides functions for density, cumulative distribution functions, quantile functions, and random generation. These functions are of the form `dxxx`, `pxxx`, `qxxx`, and `rxxx`. We are more interested in the random generation functions `rxxx`. See `help(Distributions)` for a full list. 



## Multivariate normal samples [INCOMPLETE]

Here, multivariate data refers to multiple non-independent random variables. There are two ways to generate data from a multivariate normal distribution. The first is to use a Cholesky decomposition and the second is to use `MASS::mvrnorm()`. Both approaches require a population covariance matrix.

```{r}
library(MASS)
set.seed(12479)

nobs <- 1e4

# population covariance matrix
Sigma <- matrix(c(
  3, 1, 0,
  1, 5, 0,
  0, 0, 10
), nrow = 3)

# Using chol()
x0 <- replicate(3, rnorm(nobs))
x <- (x0 %*% chol(Sigma)) + rep(c(8, 3, 2), each = nobs)

colMeans(x)

cov(x)


# Using MASS::mvrnorm()
x <- mvrnorm(n = nobs, mu = c(8, 3, 2), Sigma = Sigma)

colMeans(x)

cov(x)
```

The sample covariance matrix is roughly equivalent to the population covariance matrix `Sigma` as expected. The sample mean vector is also roughly equivalent to the population mean vector as expected. It appears that `MASS::mvrnorm` is simpler than the `chol` method. The `chol` method, however, will be useful for generating normally distributed data with additional skew and kurtosis.




## Exercises

Use `set.seed()` for these exercises for reproducible results.

1. Suppose we have $X_1, \dots, X_{20}$ such that $X_i \sim Bernoulli(p_i)$ and $p_i \sim Beta(2, 3)$. Note that the Bernoulli distribution results in 0 or 1 and the $p$ parameter ranges between 0 and 1. Meanwhile, the Beta distribution ranges between 0 and 1. Generate the random sample $X_i, \dots, X_{20}$. Hint: The Bernoulli distribution is a special case of the Binomial distribution, so `rbinom()` may be used.

1. Generate a random sample $X_1, \dots X_{100}$ such that $X_i \sim N(10, 3)$. Then, randomly assign data in this sample to be missing such that each case has a 5% chance of being missing. Hint: `sample()`, `runif()`, or `rbinom()` may be used for solving the missingness portion. Hint 2: Remember logical indexing.

1. Randomly generate a sample of size 100 with three variables that have the following characteristics. $X$ has variance 5, $Y$ has variance 3, $Z$ has variance 2, $X$ and $Y$ are correlated .3, and $Z$ is uncorrelated with the other variables. All variables have means of 0. Confirm your generation with `cov()` and `cor()`.