

# Random sampling and variate generation {#random}

```{r, include = FALSE}
set.seed(12479)
```


## Random sampling [INCOMPLETE]

`sample()`

## Univariate random samples [INCOMPLETE]

For many distributions, R provides functions for density, cumulative distribution functions, quantile functions, and random generation. These functions are of the form `dxxx`, `pxxx`, `qxxx`, and `rxxx`. We are more interested in the random generation functions `rxxx`. See `help(Distributions)` for a full list. 



## Multivariate normal samples [INCOMPLETE]

Here, multivariate data refers to multiple non-independent random variables. There are two ways to generate data from a multivariate normal distribution. The first is to use a Cholesky decomposition and the second is to use `MASS::mvrnorm()`. Both approaches require a population covariance matrix.

```{r}
library(MASS)
set.seed(12479)

nobs <- 1e4

# population covariance matrix
Sigma <- matrix(c(
  3, 1, 0,
  1, 5, 0,
  0, 0, 10
), nrow = 3)

# Using chol()
x0 <- replicate(3, rnorm(nobs))
x <- (x0 %*% chol(Sigma)) + rep(c(8, 3, 2), each = nobs)

colMeans(x)

cov(x)


# Using MASS::mvrnorm()
x <- mvrnorm(n = nobs, mu = c(8, 3, 2), Sigma = Sigma)

colMeans(x)

cov(x)
```

The sample covariance matrix is roughly equivalent to the population covariance matrix `Sigma` as expected. The sample mean vector is also roughly equivalent to the population mean vector as expected. It appears that `MASS::mvrnorm` is simpler than the `chol` method. The `chol` method, however, will be useful for generating normally distributed data with additional skew and kurtosis.


## Pseudo-randomness [INCOMPLETE]

Computer programs used for random sample generation do not produce truly random values. Instead, complex algorithms mimic randomness to produce *pseudo-random numbers*. These algorithms begin with an arbitrary starting value, or *seed*, that updates with each generated values. Unless otherwise specified, the seed is based on the internal clock of the computer's processor. (need to fact check that statement)

Pseudo-randomness for simulation studies has the advantage of reproducibility. If one writes a program and provides a fixed value for the seed, then the program will return the same results on every run. 


## Exercises

